---
title: "Geospatial Analysis Project: Clean Crime Data"
author: "Varvara Ilyina, Manjiri Satam, Gayatri Shejwal"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    code_folding: show
    df_print: paged
    highlight: tango
    number_sections: no
    theme: lumen
    toc: yes
    toc_depth: 5
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '4'
---
<style>
div.answer {background-color:#f3f0ff; border-radius: 5px; padding: 20px;}
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      eval = TRUE,
                      error = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      comment = NA,
                      fig.width = 7,
                      fig.height = 4)

```
<br>

The following R Markdown output depicts the code used to produce the data preparation and pre-processing steps for the crime data.

Specifically, we look at crime data in Northern Ireland from April 2023 until March 2024, as well as for the whole UK from March 2022 until February 2025.

<hr>

### Setup
```{r}

# clear environment
rm(list = ls())

# load libraries
pacman::p_load(tidyverse, janitor)

```

<br>

### Merge data for Northern Ireland

```{r}

# define start and end dates
start_date <- ymd("2022-04-01")
end_date <- ymd("2023-03-01")

# vector of all dates
dates <- seq.Date(from = start_date, to = end_date, by = "month") %>% 
  format("%Y-%m")

```

```{r}

## load all "street.csv" files

# loop to merge police precincts by month
for (date in dates) {
  
  # set path to the directory with .csv files for the month
  path <- paste0("./data/in/northern_ireland/", date)
  
  # list all .csv files with 'street' in their names
  files <- list.files(path, pattern = "*street.csv", full.names = TRUE)
  
  # store all .csv files in a list
  list_of_dfs <- lapply(files, read_csv)
  
  # combine all data frames into one
  if (length(list_of_dfs) > 0) {
    df_all <- bind_rows(list_of_dfs)
    
    # save the combined data frame with a name that includes the date
    file_name <- paste0("./data/out/df_", gsub("-", "_", date), ".csv")
    write_csv(df_all, file_name)
    
  } else {
    print(paste("No .csv files found for", date))
  }
  
}

```

```{r}

# define date range
dates <- seq.Date(ymd("2022-04-01"), ymd("2023-03-01"), by = "month") %>%
  format("%Y_%m")  # match file name format

file_paths <- paste0("./data/out/df_", dates, ".csv")

# load and combine all monthly files
df_crime_apr22_mar23 <- lapply(file_paths, read_csv) %>%
  bind_rows() %>%
  clean_names() %>%
  filter(!is.na(longitude), !is.na(latitude), !is.na(crime_type)) %>%
  select(month, falls_within, longitude, latitude, location, crime_type)

# save as .csv
write_csv(df_crime_apr22_mar23, "./data/out/crime_NI_apr2022_mar2023.csv")

```

<br>

### Merge data for UK

```{r}

# define start and end dates
start_date <- ymd("2022-03-01")
end_date <- ymd("2025-02-01")

# vector of all dates
dates <- seq.Date(from = start_date, to = end_date, by = "month") %>% 
  format("%Y-%m")

```

```{r}

## load all "street.csv" files

# loop to merge police precincts by month
for (date in dates) {
  
  # set path to the directory with .csv files for the month
  path <- paste0("./data/in/crime_data/", date)
  
  # list all .csv files with 'street' in their names
  files <- list.files(path, pattern = "*street.csv", full.names = TRUE)
  
  # store all .csv files in a list
  list_of_dfs <- lapply(files, read_csv)
  
  # combine all data frames into one
  if (length(list_of_dfs) > 0) {
    df_all <- bind_rows(list_of_dfs)
    
    # save the combined data frame with a name that includes the date
    file_name <- paste0("./data/out/df_", gsub("-", "_", date), ".csv")
    write_csv(df_all, file_name)
    
  } else {
    print(paste("No .csv files found for", date))
  }
  
}

```

```{r}

# define yearly spans
years <- 2022:2024

# loop to merge dfs for each year
for (yr in years) {
  
  # determine the start and end year
  start_yr <- yr
  end_yr <- yr + 1
  
  # vector of file names from march to february
  span_dates <- seq.Date(from = ymd(paste0(start_yr, "-03-01")), to = ymd(paste0(end_yr, "-02-01")), by = "month") %>%
    format("%Y-%m") %>%
    sapply(function(x) paste0("./data/out/df_", gsub("-", "_", x), ".csv"))
  
  # load all dfs for the current span
  yearly_data <- lapply(span_dates, read_csv) %>%
    bind_rows()
  
  # clean df
  df_clean <- yearly_data %>%
    janitor::clean_names() %>%
    filter(!is.na(longitude) & !is.na(latitude) & !is.na(crime_type))
  
  # save yearly data
  write_csv(df_clean, paste0("./data/out/df_", start_yr, "_", end_yr, ".csv"))
}

```
