---
title: "Geospatial Analysis Project: Visualize Crime Data"
author: "Varvara Ilyina, Manjiri Satam, Gayatri Shejwal"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    code_folding: show
    df_print: paged
    highlight: tango
    number_sections: no
    theme: lumen
    toc: yes
    toc_depth: 5
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '4'
---
<style>
div.answer {background-color:#f3f0ff; border-radius: 5px; padding: 20px;}
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      eval = TRUE,
                      error = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      comment = NA,
                      fig.width = 7,
                      fig.height = 4)

```
<br>

The following R Markdown output depicts the code used to produce the visualizations for the crime data.

Specifically, we look at crime data in Northern Ireland from April 2023 until March 2024, as well as for the whole UK from March 2022 until February 2025.

<hr>

### Setup
```{r}

# clear environment
rm(list = ls())

# load libraries
pacman::p_load(tidyverse, lubridate, spdep, sf, tmap, ggplot2, RColorBrewer)

# set tmap mode
#tmap_mode("view")

# set wd
#setwd("")

# load clean crime data
df_2022_2023 <- read_csv("./data/out/df_2022_2023.csv")
df_2023_2024 <- read_csv("./data/out/df_2023_2024.csv")
df_2024_2025 <- read_csv("./data/out/df_2024_2025.csv")

# merge data
df_all <- bind_rows(df_2022_2023, df_2023_2024, df_2024_2025) %>%
  select(month, falls_within, longitude, latitude, location, lsoa_code, lsoa_name, crime_type)

```

```{r}

# load data for northern ireland
df_crime_ni <- read_csv("./data/out/crime_NI_apr2022_mar2023.csv")

# filter UK data for april 2022 until march 2023
df_crime_uk <- df_all %>%
  filter(month >= "2022-04" & month <= "2023-03")

# save as .csv
#write_csv(df_crime_uk, "./data/out/crime_UK_apr2022_mar2023.csv")

```

<br>
<hr>
<br>

### Data Visualizations

#### Distribution of crime types

```{r}

# UK 2022 - 2025
ggplot(df_all, aes(crime_type)) +
  geom_bar() +
  labs(title = "Distribution of Crime Types",
       subtitle = "UK Crime Data, March 2022 to Feb. 2025",
       x = "Crime Type",
       y = "Count") +
  theme_classic() +
  theme(plot.title = element_text(size = 13),
        axis.title.y = element_text(size = 10),
        axis.title.x = element_text(size = 10),
        axis.text.y = element_text(size = 9),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 9))


# NI 2022 - 2023
ggplot(df_crime_ni, aes(crime_type)) +
  geom_bar() +
  labs(title = "Distribution of Crime Types",
       subtitle = "NI Crime Data, April 2022 to March 2023",
       x = "Crime Type",
       y = "Count") +
  theme_classic() +
  theme(plot.title = element_text(size = 13),
        axis.title.y = element_text(size = 10),
        axis.title.x = element_text(size = 10),
        axis.text.y = element_text(size = 9),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 9))

```

<br>

```{r}

# label region based on 'falls_within'
df_labeled <- df_2022_2023 %>%
  mutate(region = if_else(falls_within == "Police Service of Northern Ireland", 
                          "Northern Ireland", 
                          "Rest of UK"))

# summarize crime counts by region and type
df_summary <- df_labeled %>%
  count(crime_type, region) %>%
  group_by(crime_type) %>%
  mutate(prop = n / sum(n))  # proportion per crime type

# plot
ggplot(df_summary, aes(x = crime_type, y = n, fill = region)) +
  geom_col(position = "stack") +
  geom_text(aes(label = scales::percent(prop, accuracy = 1)), 
            position = position_stack(vjust = 0.8), size = 1.5, color = "black") +
  labs(title = "Crime Type Distribution: Northern Ireland vs. Rest of UK",
       subtitle = "April 2022 – March 2023",
       x = "Crime Type",
       y = "Crime Count",
       fill = "Region") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
        plot.title = element_text(size = 13),
        axis.title = element_text(size = 10),
        axis.text.y = element_text(size = 9))

```

<br>

#### Crime type by month

```{r}

# UK 2022 - 2023
df_filtered <- df_crime_uk %>%
  mutate(month = ym(month)) %>%
  mutate(month = factor(format(month, "%Y-%m"),
                        levels = format(seq(ymd("2022-04-01"), ymd("2023-03-01"), by = "1 month"), "%Y-%m")))

ggplot(df_filtered, aes(x = month, fill = crime_type)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Share of Crime Types by Month",
       subtitle = "UK Crime Data, April 2022 to March 2023",
       x = "Month",
       y = "Share of Crimes",
       fill = "Crime Type") +
  scale_fill_manual(values = colorRampPalette(brewer.pal(8, "Spectral"))(14)) +
  theme_classic() +
  theme(plot.title = element_text(size = 13),
        plot.subtitle = element_text(size = 11),
        axis.title.y = element_text(size = 10),
        axis.title.x = element_text(size = 10),
        axis.text.y = element_text(size = 9),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 9))

# save
ggsave("plots/crime_stacked_uk.png", width = 10, height = 8, dpi = 300)

```

```{r}

# NI 2022 - 2023
ggplot(df_crime_ni, aes(x = month, fill = crime_type)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Share of Crime Types by Month",
       subtitle = "NI Crime Data, April 2022 to March 2023",
       x = "Month",
       y = "Share of Crimes",
       fill = "Crime Type") +
  scale_fill_manual(values = colorRampPalette(brewer.pal(8, "Spectral"))(14)) +
  theme_classic() +
  theme(plot.title = element_text(size = 13),
        plot.subtitle = element_text(size = 11),
        axis.title.y = element_text(size = 10),
        axis.title.x = element_text(size = 10),
        axis.text.y = element_text(size = 9),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 9))

# save
ggsave("plots/crime_stacked_ni.png", width = 10, height = 8, dpi = 300)

```

<br>

#### Plot temporal trends

UK, April 2022 - March 2023
```{r}

# plot crime over time
df_plot <- df_crime_uk %>%
  mutate(month = as.Date(paste0(month, "-01"))) %>%
  group_by(month, crime_type) %>%
  summarise(n = n(), .groups = "drop") %>%
  ungroup()

# plot
ggplot(df_plot, aes(x = month,
                    y = n,
                    color = crime_type)) +
  geom_line(size = 1) +
  scale_color_viridis_d(option = "D") +
  theme_classic() +
  labs(title = "Crime trends over time by type of crime",
       subtitle = "Across the UK for April 2022 until March 2023",
       x = "Month",
       y = "Number of crimes",
       color = "Crime type")

# save
ggsave("plots/crime_acrosstime_uk.png", width = 10, height = 8, dpi = 300)

```

UK, March 2022 - February 2025
```{r}

# plot crime over time
df_plot <- df_all %>%
  mutate(month = as.Date(paste0(month, "-01"))) %>%
  group_by(month, crime_type) %>%
  summarise(n = n(), .groups = "drop") %>%
  ungroup()

# plot
ggplot(df_plot, aes(x = month,
                    y = n,
                    color = crime_type)) +
  geom_line(size = 1) +
  scale_color_viridis_d(option = "D") +
  theme_classic() +
  labs(title = "Crime trends over time by type of crime",
       subtitle = "Across the UK for March 2022 until Feb. 2025",
       x = "Month",
       y = "Number of crimes",
       color = "Crime type")

# save
ggsave("plots/crime_acrosstime_all_uk.png", width = 10, height = 8, dpi = 300)

```

<br>

### Map Crime across UK and NI

```{r}

# transform data
crime_sample <- df_crime_uk %>%
  sample_n(1000000)

# WGS 84
crime_sf <- st_as_sf(crime_sample, 
                     coords = c("longitude", "latitude"), 
                     crs = 4326)

# plot
tm_shape(crime_sf) +
  tm_dots(size = 0.05)

```

```{r}

# transform data
crime_sample <- df_crime_ni %>%
  sample_n(150000)

# WGS 84
crime_sf <- st_as_sf(crime_sample, 
                     coords = c("longitude", "latitude"), 
                     crs = 4326)

# plot
tm_shape(crime_sf) +
  tm_dots(size = 0.05)

```



