---
title: "Geospatial Analysis Project: Analyze Crime Data"
author: "Varvara Ilyina, Manjiri Satam, Gayatri Shejwal"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    code_folding: show
    df_print: paged
    highlight: tango
    number_sections: no
    theme: lumen
    toc: yes
    toc_depth: 5
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '4'
---
<style>
div.answer {background-color:#f3f0ff; border-radius: 5px; padding: 20px;}
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      eval = TRUE,
                      error = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      comment = NA,
                      fig.width = 7,
                      fig.height = 4)

```
<br>

The following R Markdown output depicts the code used to produce the Moran's I analysis for the crime data.

Specifically, we look at crime data in Northern Ireland from April 2023 until March 2024.

<hr>

### Setup
```{r}

# clear environment
rm(list = ls())

# load libraries
pacman::p_load(tidyverse, lubridate, spdep, sf, tmap, ggplot2, RColorBrewer, scales)

# set tmap mode
tmap_mode("plot")

```

<br>
<hr>
<br>

### Data Pre-processing

Load data
```{r}

# load data for northern ireland
df_crime_ni <- read_csv("./data/out/crime_NI_apr2022_mar2023.csv")

# load police boundary data
df_geo_ni <- st_read("data/in/northern_ireland_police.geojson") %>%
  mutate(zone_id = paste0("zone_", row_number()))

```

Convert crime data to a spatial object (EPSG:4326)
```{r}

df_crime_ni <- df_crime_ni %>% 
  mutate(longitude = as.numeric(longitude), latitude = as.numeric(latitude)) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>% 
  mutate(longitude = st_coordinates(.)[, 1], latitude = st_coordinates(.)[, 2])

```

Overlay crime points and police boundary polygons
```{r}

sf_use_s2(FALSE)
df_geo_ni <- st_zm(df_geo_ni) %>% st_make_valid()

df_geo_ni <- st_zm(df_geo_ni)
df_crime_ni <- st_transform(df_crime_ni, st_crs(df_geo_ni))

```

Plot
```{r}

ggplot() +
  geom_sf(data = df_geo_ni,
          fill = "white",
          color = "grey",
          alpha = 0.4) +
  geom_sf(data = df_crime_ni,
          size = 0.3,
          alpha = 0.7) +
  theme_minimal() +
  labs(title = "Crime locations in Northern Ireland",
       subtitle = "March 2022 - Feb. 2025")

# save
ggsave("plots/crime_loc_ni.png", width = 10, height = 8, dpi = 300)

```

<br>
<hr>
<br>

### Data Analysis: Moran's I

#### Use police precinct boundaries

Aggregate number of crimes occurred per polygon
```{r}

# join each crime to its containing polygon
df_crime_geo_ni <- st_join(df_crime_ni, df_geo_ni, join = st_within)

# aggregate crime counts per polygon
crime_counts <- df_crime_geo_ni %>%
  st_drop_geometry() %>%
  group_by(zone_id) %>%
  summarise(crime_count = n(), .groups = "drop")

# merge with polygons
df_crime_geo <- df_geo_ni %>%
  left_join(crime_counts, by = "zone_id")

# fill missing values with 0
df_crime_geo$crime_count[is.na(df_crime_geo$crime_count)] <- 0

```

#### Plot cloropleth + crime points

Overlay crime locations with cloropleth plot
```{r}

ggplot() +
  geom_sf(data = df_crime_geo, aes(fill = crime_count), color = "white") +
  scale_fill_distiller(
    palette = "Reds",
    direction = 1,
    name = "Total Crimes",
    breaks = c(0, 5000, 10000, 15000, 20000),
    labels = c("0–5,000", "5,000–10,000", "10,000–15,000", "15,000–20,000", "20,000+")
  ) +
  geom_sf(data = df_crime_ni, color = "darkred", size = 0.2, alpha = 0.5) +
  labs(
    title = "Crime Counts in Northern Ireland",
    subtitle = "April 2022 – March 2023"
  ) +
  theme_minimal() +
  theme(
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8),
    legend.key.height = unit(0.4, "cm"),
    legend.key.width = unit(0.6, "cm")
  )

# save plot
ggsave("plots/crime_loc_cloropleth_ni.png", width = 10, height = 8, dpi = 300)

```
How to interpret:

- Cloropleth plot used to identify hotpots of crime
- Color gradient: The darker the red, the more crimes within that police precinct
- Top crime areas: Dark red areas, such as Belfast and Derry, may correspond to urban centers with more population density and indicate higher total crime volumes

<br>

#### Calculate Moran's I
```{r}

# create spatial neighbors
nb <- poly2nb(df_crime_geo)
lw <- nb2listw(nb, style = "W", zero.policy = TRUE)

# run Local Moran's I
local_moran <- localmoran(df_crime_geo$crime_count, lw, zero.policy = TRUE)

# add results
df_crime_geo$local_I <- local_moran[, "Ii"]
df_crime_geo$p_val <- local_moran[, "Pr(z != E(Ii))"]

```

Classify into LISA quadrants
```{r}

# compute global mean and lagged mean
mean_crime <- mean(df_crime_geo$crime_count, na.rm = TRUE)
lagged_crime <- lag.listw(lw, df_crime_geo$crime_count)

df_crime_geo$quadrant <- case_when(
  df_crime_geo$crime_count > mean_crime & lagged_crime > mean_crime & df_crime_geo$p_val < 0.05 ~ "High-High",
  df_crime_geo$crime_count < mean_crime & lagged_crime < mean_crime & df_crime_geo$p_val < 0.05 ~ "Low-Low",
  df_crime_geo$crime_count > mean_crime & lagged_crime < mean_crime & df_crime_geo$p_val < 0.05 ~ "High-Low",
  df_crime_geo$crime_count < mean_crime & lagged_crime > mean_crime & df_crime_geo$p_val < 0.05 ~ "Low-High",
  TRUE ~ "Non-significant"
)

```

Plot LISA map
```{r}

tm_shape(df_crime_geo) +
  tm_polygons("quadrant",
              palette = c(
                "High-High" = "darkred",
                "Low-Low" = "steelblue",
                "High-Low" = "pink",
                "Low-High" = "skyblue",
                "Non-significant" = "white"
              ),
              title = "Local Moran's I Quadrants") +
  tm_layout(legend.outside = TRUE)

```

```{r}

# save to .png
tmap_save(
  tm = tm_shape(df_crime_geo) +
    tm_polygons("quadrant",
                palette = c(
                  "High-High" = "darkred",
                  "Low-Low" = "steelblue",
                  "High-Low" = "pink",
                  "Low-High" = "skyblue",
                  "Non-significant" = "white"
                ),
                title = "Local Moran's I Quadrants") +
    tm_layout(legend.outside = TRUE),
  filename = "plots/lisa_map_ni_geo.png",
  width = 2000,
  height = 1600,
  units = "px",
  dpi = 300
)

```

How to interpret:

- High-High (dark red): Areas with high crime surrounded by other high-crime areas, pointing to crime hotspots
- Low-High (light blue): Areas with low crime surrounded by high-crime areas, possibly indicating under-policed or isolated areas
- Non-significant (white): No clear spatial pattern found, could be random

- High-High cluster in and around Belfast indicates a crime center surrounded by similar areas
- the Low-High area next to Belfast may point to a suburban region that surrounds the urban center with more crime

<br>
<hr>
<br>

#### Use Super Output Areas (SOA)

Granular data downloaded from: https://www.nisra.gov.uk/support/output-geography-census-2011/super-output-areas-census-2011?utm_source=chatgpt.com#toc-2

Northern Ireland doesn't use LSOAs like England or Wales, but similar fine units include Super Output Areas (SOAs) and Small Areas (SAs). Northern Ireland contains 890 SOAs, each with about 2000 people per unit. SAs are even smaller, containing only about 125 households.

<hr>

Load shapefile with SOA boundary polygons
```{r}

df_soa_ni <- st_read("data/in/SOA2011_esri_Shapefile/SOA2011.shp") %>%
  st_zm() %>%
  mutate(soa_id = paste0("SOA_", row_number()))

```

Convert crime data to sf points and match CRS
```{r}

# transform crime data to match SOA CRS
df_crime_ni <- st_transform(df_crime_ni, st_crs(df_soa_ni))

```

Join crimes to SOA areas
```{r}

df_crime_soa_ni <- st_join(df_crime_ni, df_soa_ni, join = st_within)

```

Aggregate crimes by SOA
```{r}

crime_counts <- df_crime_soa_ni %>%
  st_drop_geometry() %>%
  group_by(soa_id) %>%
  summarise(crime_count = n(), .groups = "drop")

df_crime_soa <- df_soa_ni %>%
  left_join(crime_counts, by = "soa_id") %>%
  mutate(crime_count = ifelse(is.na(crime_count), 0, crime_count))

```

#### Plot cloropleth
```{r, warning=FALSE}

# plot
tm_shape(df_crime_soa) +
  tm_polygons("crime_count",
              palette = "brewer.reds",
              title = "Total Crimes") +
  tm_layout(
    main.title = "Crime Counts by Super Output Area\nApril 2022 – March 2023",
    main.title.size = 1.2,
    main.title.position = "center",
    legend.outside = TRUE
  )

# save cloropleth
tmap_save(
  filename = "plots/crime_choropleth_ni_soa.png",
  width = 2000, height = 1600, dpi = 300
)

```

#### Calculate Moran's I
```{r}

nb <- poly2nb(df_crime_soa)
lw <- nb2listw(nb, style = "W", zero.policy = TRUE)
local_moran <- localmoran(df_crime_soa$crime_count, lw, zero.policy = TRUE)

df_crime_soa$local_I <- local_moran[, "Ii"]
df_crime_soa$p_val   <- local_moran[, "Pr(z != E(Ii))"]

mean_crime <- mean(df_crime_soa$crime_count, na.rm = TRUE)
lagged_crime <- lag.listw(lw, df_crime_soa$crime_count)

df_crime_soa$quadrant <- case_when(
  df_crime_soa$crime_count > mean_crime & lagged_crime > mean_crime & df_crime_soa$p_val < 0.05 ~ "High-High",
  df_crime_soa$crime_count < mean_crime & lagged_crime < mean_crime & df_crime_soa$p_val < 0.05 ~ "Low-Low",
  df_crime_soa$crime_count > mean_crime & lagged_crime < mean_crime & df_crime_soa$p_val < 0.05 ~ "High-Low",
  df_crime_soa$crime_count < mean_crime & lagged_crime > mean_crime & df_crime_soa$p_val < 0.05 ~ "Low-High",
  TRUE ~ "Non-significant"
)

```

Plot LISA map with quadrants
```{r}

tm_shape(df_crime_soa) +
  tm_polygons("quadrant",
              palette = c(
                "High-High" = "darkred",
                "Low-Low" = "steelblue",
                "High-Low" = "pink",
                "Low-High" = "skyblue",
                "Non-significant" = "white"
              ),
              title = "Local Moran's I Quadrants") +
  tm_layout(legend.outside = TRUE)

```



